\thornado-hydro: towards discontinuous galerkin methods for supernova hydrodynamics

### 2018-10-12 Documents/Euler/Astronum_2018/manuscript.tex
First full draft of paper

\begin{document}
\title{\thornado-hydro: towards discontinuous galerkin methods for supernova hydrodynamics}

\author{Eirik Endeve}
\address{Computational \& Applied Mathematics, Oak Ridge National Laboratory, Oak Ridge, TN 37831 USA}
\address{Department of Physics and Astronomy, University of Tennessee Knoxville, TN 37996-1200}
\ead{endevee@ornl.gov}

\author{Kristopher Andrew, Brandon Barker, Jesse Buffaloe, Samuel Dunham, David Pochik, Juliana Pulsinelli, Nick Roberts}

\begin{abstract}
The toolkit for high-order neutrino-radiation hydrodynamics (\thornado) is being developed for simulations of core-collapse supernovae (CCSN) and related problems.  
Current capabilities in \thornado\ include solvers for the Euler equations --- in non-relativistic and special relativistic limits --- and the two-moment model of neutrino transport.  
The spatial discretization in \thornado\ is based on the discontinuous Galerkin (DG) method, which is receiving increased attention from the computational astrophysics community.  
In this paper, we provide an overview of the numerical methods for the Euler equations in \thornado, and present some encouraging preliminary numerical results from a set of basic tests in one and two spatial dimensions.  
\end{abstract}

@@

\ee{Introduction coming soon.}

@@

Excellent books and review articles on the DG method are available (see, e.g., \cite{cockburnShu_1998,cockburnShu_2001,hesthavenWarburton_2008,shu_2016} and references therein), and we will not go into too much detail here.  
However, we review some key concepts to introduce notation, and emphasize specific choices in our implementation in \thornado.  
To this end, we consider a system of conservation laws with sources of the form
\begin{equation}
  \pd{}{t}\big(\,\sqrt{\gamma}\,\vect{U}\,\big)
  +\pd{}{i}\Big(\,\sqrt{\gamma}\,\vect{F}^{i}(\vect{U})\,\Big)
  =\sqrt{\gamma}\,\vect{S}(\vect{U}),
  \label{eq:extendedEulerCompact}
\end{equation}
where $\vect{U}$ is the evolved state vector, $\vect{F}^{i}$ the fluxes, and $\vect{S}$ the source vector.  
We use a formulation of the equations sufficiently general to accommodate curvilinear spatial coordinates encoded in the metric $\gamma_{ij}$, giving the squared line element $ds^{2}=\gamma_{ij}dx^{i}dx^{j}$, whose determinant is $\gamma$.  
To solve Eq.~\eqref{eq:extendedEulerCompact}, the computational domain $D$ is divided into a disjoint union $\cT$ of open elements $\bK$, so that $D = \cup_{\bK \in \cT}\bK$.  
We require that each element is a box in the logical coordinates; i.e.,
\begin{equation}
  \bK=\{\vect{x} : x^{i} \in K^{i} := (\xL^{i},\xH^{i})\}, 
\end{equation}
with the surface elements denoted $\partial\bK^{i}=\times_{j\ne i}K^{j}$.  
We let $V_{\bK}$ denote the proper element volume
\begin{equation}
  V_{\bK} = \int_{\bK}dV, \quad\text{where}\quad dV = \sqrt{\gamma}\,\prod_{i=1}^{d}dx^{i}.  
\end{equation}
We also define as a set $\bx=\{\tilde{\bx}^{i},x^{i}\}$, and $\dx^{i}=\xH^{i}-\xL^{i}$.

We let the approximation space for the DG method, $\mathbb{V}^{k}$, be constructed from the tensor product of one-dimensional polynomials of maximal degree $k$.  
(Functions in $\mathbb{V}^{k}$ can be discontinuous across element interfaces.)  
The semi-discrete DG problem is to find $\vect{U}_{h}\in\mathbb{V}^{k}$ (which approximates $\vect{U}$ in Eq.~\eqref{eq:extendedEulerCompact}) such that for all $v\in\mathbb{V}^{k}$ and all $\bK\in\mathcal{T}$
\begin{align}
  &\partial_{t}\int_{\bK}\vect{U}_{h}\,v\,dV
  +\sum_{i=1}^{d}\int_{\partial\bK^{i}}\big(\,\sqrt{\gamma}\,\widehat{\vect{F}}^{i}(\vect{U}_{h})\,v\big|_{\xH^{i}}-\sqrt{\gamma}\,\widehat{\vect{F}}^{i}(\vect{U}_{h})\,v\big|_{\xL^{i}}\,\big)\,d\tilde{\bx}^{i} \nonumber \\
  &\hspace{24pt}
  -\sum_{i=1}^{d}\int_{\bK}\vect{F}^{i}(\vect{U}_{h})\,\pd{v}{i}\,dV
  =\int_{\bK}\vect{S}(\vect{U}_{h})\,v\,dV.  
  \label{eq:semidiscreteDG}
\end{align}
In Eq.~\eqref{eq:semidiscreteDG}, $\widehat{\vect{F}}^{i}(\vect{U}_{h})$ is a numerical flux approximating the flux on the $i$th surface of $\bK$.  
The numerical flux function $\vect{f}^{i}$ evaluates the numerical flux given values from both sides of an element interface; i.e.,
\begin{equation}
  \widehat{\vect{F}}^{i}(\vect{U}_{h})=\vect{f}^{i}(\vect{U}_{h}(x^{i,-},\tilde{\bx}^{i}),\vect{U}_{h}(x^{i,+},\tilde{\bx}^{i})),
\end{equation}
where superscripts $-/+$, e.g., in the arguments of $\vect{U}_{h}$, indicate that the function is evaluated to the immediate left/right of $x^{i}$.  
We use the Harten-Lax-van Leer (HLL) flux \cite{harten_etal_1983} or the HLLC flux \cite{toro_etal_1994,mignoneBodo_2005} for all the numerical experiments presented in Section~\ref{sec:numerical}.

We provide further details on the DG method to arrive at the equations that are actually evolved in \thornado.  
We start by introducing some notation, defining the polynomial expansion for $\vect{U}_{h}$, and the quadrature rules used to evaluate the integrals in Equation~\eqref{eq:semidiscreteDG}.  
Then we provide explicit expressions for each of the terms in Equation~\eqref{eq:semidiscreteDG}.  
For simplicity we consider one spatial dimension ($d=1$) and drop indices denoting the spatial dimension.  
In each element $K$, we use a nodal representation of the conserved variables $\vect{U}$; i.e.,
\begin{equation}
  \vect{U}(x,t)\approx
  \vect{U}_{h}(x,t)=\sum_{i=1}^{N}\vect{U}_{i}(t)\,\ell_{i}(x),
  \label{eq:conservedNodalExpansion}
\end{equation}
where $\ell_{i}(x)$ are Lagrange polynomials
\begin{equation}
  \ell_{i}(\eta)=
  \prod_{\substack{j=1\\j\ne i}}^{N}\f{\eta-\eta_{j}}{\eta_{i}-\eta_{j}},
\end{equation}
defined on the reference element $I = \{ \eta : \eta \in (-0.5,0.5) \}$, and constructed to interpolate the node set $S_{N}=\{\eta_{i}\}_{i=1}^{N}\subset I$.  
The spatial coordinate $x$ and the reference coordinate $\eta$ are related by the mapping $x(\eta)=\xL+(0.5+\eta)\,\dx$.  
Then, for any $\eta_{j}\in S_{N}$, $\ell_{i}(\eta_{j})=\delta_{ij}$, so that $\vect{U}_{h}(x(\eta_{j}),t)=\vect{U}_{j}(t)$.  
We introduce numerical quadratures to evaluate the integrals in Equation~\eqref{eq:semidiscreteDG}.  
First we define the $M$-point quadrature $Q_{M}:C^{0}(I)\to\mathbb{R}$ with abscissas $\hat{S}_{M}=\{\eta_{q}\}_{q=1}^{M}$ and weights $\{w_{q}\}_{q=1}^{M}$, normalized such that $\sum_{q=1}^{M}w_{q}=1$.  
The $M$-point Legendre-Gauss quadrature, which we use, integrates polynomials of degree $\le 2M-1$ exactly.  
Then, if $P(x)$ is such a polynomial, we have
\begin{equation}
  \f{1}{\dx}\int_{K}P(x)\,dx=\int_{I}P(\eta)\,d\eta=Q_{N}\big[P\big]\equiv\sum_{q=1}^{M}w_{q}\,P(\eta_{q}).  
\end{equation}

Note that the interpolation points $S_{N}$ and the quadrature points $\hat{S}_{M}$ generally do not coincide.  
However, for sake of computational efficiency we let $M=N$ and $S_{N}=\hat{S}_{N}$, which is a spectral-type nodal collocation DG approximation \cite{bassi_etal_2013}.  
Inserting Eq.~\eqref{eq:conservedNodalExpansion} into Eq.~\eqref{eq:semidiscreteDG}, letting $v(x)=\ell_{k}(x)$, and using the quadratures defined above, we obtain
\begin{align}
  \pd{}{t}\int_{K}\vect{U}_{h}\,v\,dV
  &\approx w_{k}\,\sqrt{\gamma}_{k}\,\pd{}{t}\vect{U}_{k}\,\dx
  \label{eq:timeDerivativeTerm}
\end{align}
for the time derivative term.  
The integration is approximate since we use the Legendre-Gauss quadrature rule with the nodal points given by the expansion in Eq.~\eqref{eq:conservedNodalExpansion}.  
Since $\ell_{i}(\eta_{k})=\delta_{ik}$, this leads to a diagonal mass matrix, and simplifies the implementation.  
Similarly, we obtain
\begin{align}
  \int_{K}\vect{S}(\vect{U}_{h})\,v\,dV
  &\approx w_{k}\,\sqrt{\gamma}_{k}\,\vect{S}(\vect{U}_{k})\,\dx.
  \label{eq:sourceTerm}
\end{align}
for the source term.  
Finally, the volume term (last term on the left-hand side of Eq.~\eqref{eq:semidiscreteDG}) becomes
\begin{equation}
  \int_{K}\vect{F}(\vect{U}_{h})\,\pderiv{v}{x}\,dV
  \approx \sum_{q=1}^{N}w_{q}\,\sqrt{\gamma}_{q}\,\vect{F}(\vect{U}_{q})\,\pderiv{\ell_{k}}{\eta}(\eta_{q}).
  \label{eq:volumeTerm}
\end{equation}
Using Eqs.~\eqref{eq:timeDerivativeTerm}-\eqref{eq:volumeTerm} in Eq.~\eqref{eq:semidiscreteDG} results in the semi-discrete form
\begin{align}
  \deriv{\vect{U}_{k}}{t}
  =-\f{1}{w_{k}\sqrt{\gamma}_{k}\dx}
  \Big\{
  \Big[\,
    \sqrt{\gamma}\widehat{\vect{F}}\ell_{k}\big|_{\xH}
    -\sqrt{\gamma}\widehat{\vect{F}}\ell_{k}\big|_{\xL}
  \,\Big]
  -\sum_{q=1}^{N}w_{q}\,\sqrt{\gamma}_{q}\,\vect{F}(\vect{U}_{q})\,\pderiv{\ell_{k}}{\eta}(\eta_{q})
  \Big\} + \vect{S}(\vect{U}_{k}).
  \label{eq:semidiscreteDiscretized}
\end{align}
Eq.~\eqref{eq:semidiscreteDiscretized} comprises a system of ordinary differential equations (ODEs), which are integrated in time with an ODE solver.  
In section~\ref{sec:numerical} we use the optimal third-order strong stability-preserving Runge-Kutta (SSP-RK3) method from \cite{shuOsher_1988}.

@@

We have run this test with $200$ elements to $t=1.8$.  
In the two upper panels of Figure~\ref{fig:ShuOsher} we plot the density for various values of the troubled-cell indicator threshold $C_{\TCI}$; $0.0$ (full limiting, green), $0.03$ (magenta), $0.3$ (red), and $3.0$ (blue).  
(Larger values of $C_{\TCI}$ imply less limiting.)  
The results are compared with a reference solution computed with $2048$ elements.  
With full limiting, the scheme is not able to capture the density variations behind the shock (cf. upper right panel in Figure~\ref{fig:ShuOsher}).  
When $C_{\TCI}=0.03$, the variations are resolved, but the amplitudes are significantly reduced when compared with the reference solution.  
With $C_{\TCI}=3.0$, the variations are well resolved, and the amplitudes are comparable to the reference.  
\begin{figure}[h]
  \centering
  \begin{minipage}{18pc}
    \includegraphics[width=18pc]{./Figures/ShuOsher_Astronum_2018}
  \end{minipage}\hspace{0.5pc}%
  \begin{minipage}{18pc}
    \includegraphics[width=18pc]{./Figures/ShuOsher_Inset_Astronum_2018}
  \end{minipage} \\
  \begin{minipage}{12pc}
    \includegraphics[width=12pc]{./Figures/ShuOsher_TCI_0015_Astronum_2018}
  \end{minipage}\hspace{0.5pc}%
  \begin{minipage}{12pc}
    \includegraphics[width=12pc]{./Figures/ShuOsher_TCI_0150_Astronum_2018}
  \end{minipage}\hspace{0.5pc}
  \begin{minipage}{12pc}
    \includegraphics[width=12pc]{./Figures/ShuOsher_TCI_1500_Astronum_2018}
  \end{minipage}
  \caption{\label{fig:ShuOsher}Results for the Shu-Osher shock tube problem computed with $200$ elements.  The upper panels show the mass density at $t=1.8$ for various values of the troubled-cell indicator threshold $C_{\TCI}$, compared with a high-resolution (2048 elements) reference solution.  In the lower panels we show elements flagged for limiting in the $xt$-plane for various values of $C_{\TCI}$.}
\end{figure}
In the lower panels of Figure~\ref{fig:ShuOsher} we plot locations in the $xt$-plane where elements have been flagged for limiting by the troubled-cell indicator: $C_{\TCI}=0.03$ (left), $C_{\TCI}=0.3$ (middle), and $C_{\TCI}=3.0$ (right).  
With $C_{\TCI}=0.03$, limiting is triggered in a wide region around the shock and also around the peaks of the three leftmost density variations seen in upper left panel in Figure~\ref{fig:ShuOsher}.  
With $C_{\TCI}=3.0$, limiting is only triggered in the shock.

\subsubsection{Kelvin-Helmholtz (KH) Instability}
This hydrodynamic instability is of significant astrophysical relevance --- including in the explosion phase of CCSNe.  
It may occur in the interface separating fluids in relative motion.  
We compute this problem on the unit square $D=[0,1]\times[0,1]$, imposing periodic boundary conditions, and with initial conditions and single-mode perturbation taken from \cite{mcnally_etal_2012}.  
We use $256^{2}$ elements and evolve until $t=3.0$, well into the nonlinear regime.  
Results are displayed in Figure~\ref{fig:KelvinHelmholtz}.  
We have computed two models, one with $C_{\TCI}=0.06$ (left panels) and one without limiting ($C_{\TCI}\to\infty$; right panels).

@@

This test is of immediate relevance to simulation of CCSNe.  
The initial condition is described in detail in \cite{blondin_etal_2003}, and has been used by many to study the role of the standing accretion shock instability (SASI) on CCSN explosion dynamics.  
We follow closely the description in \cite{blondin_etal_2003} to initialize this test.  
We use spherical polar coordinates, let $D=[0.2,2.0]$, and $\Gamma=4/3$.  
The gravitational potential is given by the point-mass formula with $GM=0.5$.  
A stationary shock is placed at a radius $R_{\shock}=1$.  
Ahead of the shock, the flow is essentially in free-fall towards the shock with a constant Mach number of $100$.  
The mass accretion rate is held fixed at the outer boundary to $\dot{M}=4\pi$.  
Below the shock, $r\in[0.2,1]$, the settling solution is obtained by solving Bernoulli's equation.  
We let matter flow through the inner boundary, and the density and pressure in the boundary elements are set by extrapolating from $D$ assuming $\rho\propto r^{-3}$ and $E\propto r^{-4}$, while the momentum components are held fixed to their initial values.

>>>>

\begin{document}
\title{\thornado-hydro: towards discontinuous galerkin methods for supernova hydrodynamics\footnote{This manuscript has been authored by UT-Battelle, LLC under Contract No. DE-AC05-00OR22725 with the U.S. Department of Energy. The United States Government retains and the publisher, by accepting the article for publication, acknowledges that the United States Government retains a non-exclusive, paid-up, irrevocable, world-wide license to publish or reproduce the published form of this manuscript, or allow others to do so, for United States Government purposes. The Department of Energy will provide public access to these results of federally sponsored research in accordance with the DOE Public Access Plan(http://energy.gov/downloads/doe-public-access-plan).}}

\author{Eirik Endeve$^{1,2,6}$, Jesse Buffaloe$^{2}$, Samuel Dunham$^{3}$, Nick Roberts$^{2}$, Kristopher Andrew$^{4,1}$, Brandon Barker$^{2}$, David Pochik$^{2}$, Juliana Pulsinelli$^{5,7}$ and Anthony Mezzacappa$^{2,6}$}
\address{$^{1}$Computer Science and Mathematics Division, Oak Ridge National Laboratory, TN 37831}
\address{$^{2}$Department of Physics and Astronomy, University of Tennessee Knoxville, TN 37996}
\address{$^{3}$Department of Astronomy, Vanderbilt University, TN 37212}
\address{$^{4}$Department of Physics and Astronomy, University of Kentucky Lexington, KY 40506}
\address{$^{5}$Department of Astrophysical Sciences, Princeton University, NJ 08544}
\address{$^{6}$Joint Institute for Computational Sciences, Oak Ridge National Laboratory, TN 37831}
\address{$^{7}$Physics Division, Oak Ridge National Laboratory, TN 37831}
\ead{endevee@ornl.gov}

\begin{abstract}
The toolkit for high-order neutrino-radiation hydrodynamics (\thornado) is being developed for simulations of core-collapse supernovae (CCSNe) and related problems.  
Current capabilities in \thornado\ include solvers for the Euler equations --- in non-relativistic and special relativistic limits --- and the two-moment model of neutrino transport.  
The spatial discretization in \thornado\ is based on the discontinuous Galerkin (DG) method, which is receiving increased attention from the computational astrophysics community.  
In this paper, we provide an overview of the numerical methods for the Euler equations in \thornado, and present some encouraging preliminary numerical results from a set of basic tests in one and two spatial dimensions.  
\end{abstract}

@@

Along with solvers for neutrino transport and gravity, the hydrodynamics solver constitutes a major component of core-collapse supernova (CCSN) simulations.  
The dynamics of the bounce shock and associated fluid instabilities, which drive turbulent flows and help shape the explosion, must be faithfully captured (see, e.g., \cite{muller_2016}, for a recent review).  
Essentially, the stellar interior is modeled as a perfect fluid (i.e., thermal conduction and viscosity are not explicitly included) using the Euler equations, but extended to accommodate a nuclear equation of state.

The discontinuous Galerkin (DG) method \cite{cockburnShu_2001,hesthavenWarburton_2008} appears as an appealing choice to model fluid flows in CCSNe.  
DG methods combine elements of spectral and finite volume methods, and achieve high-order accuracy on a compact stencil.  
Data is only communicated with nearest neighbors, regardless of the formal order of accuracy, which leads to a high computation to communication ratio, and favorable parallel scalability on heterogeneous architectures has been demonstrated \cite{klockner_etal_2009}.  
They can be used in combination with $hp$-adaptivity \cite{remacle_etal_2003}, where, in addition to grid refinement with AMR, the local polynomial degree can be chosen differently, and independently, in different cells.  
DG methods can easily be applied to problems involving curvilinear coordinates, which is beneficial in numerical relativity \cite{teukolsky_2016}.  
Currently, most hydrodynamics solvers in CCSN codes are based on finite volume methods.  
Exploring the utility and performance of DG methods to model CCSNe seems like a worthy exercise by itself.

The toolkit for high-order neutrino-radiation hydrodynamics (\thornado) is being developed to simulate neutrino-radiation hydrodynamics in CCSNe and related applications in nuclear astrophysics.  
The spatial discretization of solvers for hyperbolic partial differential equations in \thornado\ is based on the DG method, while we foresee using a combination of spectral and continuous finite element methods for solving elliptic equations; e.g., for Newtonian gravity or relativistic gravity employing the conformal flatness approximation \cite{wilson_etal_1996}.  
With these approaches, we employ high-order discretization techniques based on a common mathematical framework in all the major model components.  
In this paper, we provide an initial description and encouraging demonstration of the DG method implemented in \thornado\ to solve the Euler equations.  
We note that others (e.g., \cite{radiceRezzolla_2011,schaal_etal_2015}) have also investigated the DG method in the context of astrophysical hydrodynamics and reported encouraging results.  
Whether the high-order approach will improve accuracy and efficiency of CCSN models remains to be demonstrated.

@@

Excellent books and review articles on the DG method are available (see, e.g., \cite{cockburnShu_1998,cockburnShu_2001,hesthavenWarburton_2008,shu_2016} and references therein), and we will not go into too much detail here.  
However, we review some key concepts to introduce notation, and emphasize specific choices for our implementation in \thornado.  
To this end, we consider a system of conservation laws with sources of the form
\begin{equation}
  \pd{}{t}\big(\,\sqrt{\gamma}\,\vect{U}\,\big)
  +\sum_{i=1}^{d}\pd{}{i}\Big(\,\sqrt{\gamma}\,\vect{F}^{i}(\vect{U})\,\Big)
  =\sqrt{\gamma}\,\vect{S}(\vect{U}),
  \label{eq:extendedEulerCompact}
\end{equation}
where $\vect{U}$ is the evolved state vector, $\vect{F}^{i}$ the fluxes, and $\vect{S}$ the source vector.  
We use a formulation of the equations sufficiently general to accommodate curvilinear spatial coordinates encoded in the metric $\gamma_{ij}$, giving the squared line element $ds^{2}=\gamma_{ij}dx^{i}dx^{j}$, whose determinant is $\gamma$.  
To solve Eq.~\eqref{eq:extendedEulerCompact}, the computational domain $D\subset\mathbb{R}^{d}$ is divided into a disjoint union $\cT$ of open elements $\bK$, so that $D = \cup_{\bK \in \cT}\bK$.  
We require that each element is a box in the logical coordinates; i.e.,
\begin{equation}
  \bK=\{\,\vect{x} : x^{i} \in K^{i} := (\xL^{i},\xH^{i}),\,i=1,\ldots,d\,\}, 
\end{equation}
with the surface elements denoted $\partial\bK^{i}=\times_{j\ne i}K^{j}$.  
We let $V_{\bK}$ denote the proper element volume
\begin{equation}
  V_{\bK} = \int_{\bK}dV, \quad\text{where}\quad dV = \sqrt{\gamma}\,\prod_{i=1}^{d}dx^{i}.  
\end{equation}
We also define as a set $\bx=\{\tilde{\bx}^{i},x^{i}\}$ and $\dx^{i}=\xH^{i}-\xL^{i}$.

We let the approximation space for the DG method, $\mathbb{V}^{k}$, be constructed from the tensor product of one-dimensional polynomials of maximal degree $k$.  
Note that functions in $\mathbb{V}^{k}$ can be discontinuous across element interfaces.  
The semi-discrete DG problem is to find $\vect{U}_{h}\in\mathbb{V}^{k}$, which approximates $\vect{U}$ in Eq.~\eqref{eq:extendedEulerCompact}, such that for all $v\in\mathbb{V}^{k}$ and all $\bK\in\mathcal{T}$
\begin{align}
  &\partial_{t}\int_{\bK}\vect{U}_{h}\,v\,dV
  +\sum_{i=1}^{d}\int_{\partial\bK^{i}}\big(\,\sqrt{\gamma}\,\widehat{\vect{F}}^{i}(\vect{U}_{h})\,v\big|_{\xH^{i}}-\sqrt{\gamma}\,\widehat{\vect{F}}^{i}(\vect{U}_{h})\,v\big|_{\xL^{i}}\,\big)\,d\tilde{\bx}^{i} \nonumber \\
  &\hspace{24pt}
  -\sum_{i=1}^{d}\int_{\bK}\vect{F}^{i}(\vect{U}_{h})\,\pd{v}{i}\,dV
  =\int_{\bK}\vect{S}(\vect{U}_{h})\,v\,dV.  
  \label{eq:semidiscreteDG}
\end{align}
In Eq.~\eqref{eq:semidiscreteDG}, $\widehat{\vect{F}}^{i}(\vect{U}_{h})$ is a numerical flux approximating the flux on the $i$th surface of $\bK$.  
The numerical flux function $\vect{f}^{i}$ is evaluated using values from both sides of an element interface; i.e.,
\begin{equation}
  \widehat{\vect{F}}^{i}(\vect{U}_{h})=\vect{f}^{i}(\vect{U}_{h}(x^{i,-},\tilde{\bx}^{i}),\vect{U}_{h}(x^{i,+},\tilde{\bx}^{i})),
\end{equation}
where superscripts $-/+$, e.g., in the arguments of $\vect{U}_{h}$, indicate that the function is evaluated to the immediate left/right of $x^{i}$.  
We use the Harten-Lax-van Leer (HLL) flux \cite{harten_etal_1983} or the HLLC flux \cite{toro_etal_1994,mignoneBodo_2005} for all the numerical experiments presented in Section~\ref{sec:numerical}.

We provide further details on the DG method to arrive at the equations that are actually evolved in \thornado.  
We start by introducing some notation, defining the polynomial expansion for $\vect{U}_{h}$ and the quadrature rules used to evaluate the integrals in Equation~\eqref{eq:semidiscreteDG}.  
Then we provide explicit expressions for each of the terms in Equation~\eqref{eq:semidiscreteDG}.  
For simplicity we consider one spatial dimension ($d=1$) and drop indices denoting the spatial dimension.  
In each element $K$, we use a nodal representation of the conserved variables $\vect{U}$; i.e.,
\begin{equation}
  \vect{U}(x,t)\approx
  \vect{U}_{h}(x,t)=\sum_{i=1}^{N}\vect{U}_{i}(t)\,\ell_{i}(x),
  \quad\text{where}\quad
  \ell_{i}(\eta)=
  \prod_{\substack{j=1\\j\ne i}}^{N}\f{\eta-\eta_{j}}{\eta_{i}-\eta_{j}}
  \label{eq:conservedNodalExpansion}
\end{equation}
are Lagrange polynomials defined on the reference element $I = \{ \eta : \eta \in (-0.5,0.5) \}$, and constructed to interpolate the node set $S_{N}=\{\eta_{i}\}_{i=1}^{N}\subset I$.  
The spatial coordinate $x$ and the reference coordinate $\eta$ are related by the mapping $x(\eta)=\xL+(0.5+\eta)\,\dx$.  
Then, for any $\eta_{j}\in S_{N}$, $\ell_{i}(\eta_{j})=\delta_{ij}$, so that $\vect{U}_{h}(x(\eta_{j}),t)=\vect{U}_{j}(t)$.  
We introduce numerical quadratures to evaluate the integrals in Equation~\eqref{eq:semidiscreteDG}.  
First we define the $M$-point quadrature $Q_{M}:C^{0}(I)\to\mathbb{R}$ with abscissas $\hat{S}_{M}=\{\eta_{q}\}_{q=1}^{M}$ and weights $\{w_{q}\}_{q=1}^{M}$, normalized such that $\sum_{q=1}^{M}w_{q}=1$.  
The $M$-point Legendre-Gauss quadrature, which we use, integrates polynomials of degree $\le 2M-1$ exactly.  
Then, if $P_{h}(x)$ is such a polynomial, we have
\begin{equation}
  \f{1}{\dx}\int_{K}P_{h}(x)\,dx=\int_{I}P_{h}(\eta)\,d\eta=Q_{M}\big[P_{h}\big]\equiv\sum_{q=1}^{M}w_{q}\,P_{h}(\eta_{q}).  
\end{equation}

Note that the interpolation points $S_{N}$ and the quadrature points $\hat{S}_{M}$ generally do not coincide.  
However, for sake of computational efficiency we let $M=N$ and $S_{N}=\hat{S}_{N}$, which is a spectral-type nodal collocation DG approximation \cite{bassi_etal_2013}.  
Inserting Eq.~\eqref{eq:conservedNodalExpansion} into Eq.~\eqref{eq:semidiscreteDG}, letting $v(x)=\ell_{k}(x)$, and using the quadratures defined above, we obtain
\begin{align}
  \pd{}{t}\int_{K}\vect{U}_{h}\,v\,dV
  &\approx w_{k}\,\sqrt{\gamma}_{k}\,\pd{}{t}\vect{U}_{k}\,\dx
  \label{eq:timeDerivativeTerm}
\end{align}
for the time derivative term.  
For a general metric, the integral in Eq.~\eqref{eq:timeDerivativeTerm} is approximate since we use the Legendre-Gauss quadrature rule with the nodal points given by the expansion in Eq.~\eqref{eq:conservedNodalExpansion}.  
However, since $\ell_{i}(\eta_{k})=\delta_{ik}$, this leads to a diagonal mass matrix, and simplifies the implementation.  
Similarly, we obtain
\begin{align}
  \int_{K}\vect{S}(\vect{U}_{h})\,v\,dV
  &\approx w_{k}\,\sqrt{\gamma}_{k}\,\vect{S}(\vect{U}_{k})\,\dx.
  \label{eq:sourceTerm}
\end{align}
for the source term.  
Finally, the volume term (last term on the left-hand side of Eq.~\eqref{eq:semidiscreteDG}) becomes
\begin{equation}
  \int_{K}\vect{F}(\vect{U}_{h})\,\pderiv{v}{x}\,dV
  \approx \sum_{q=1}^{N}w_{q}\,\sqrt{\gamma}_{q}\,\vect{F}(\vect{U}_{q})\,\pderiv{\ell_{k}}{\eta}(\eta_{q}).
  \label{eq:volumeTerm}
\end{equation}
Using Eqs.~\eqref{eq:timeDerivativeTerm}-\eqref{eq:volumeTerm} in Eq.~\eqref{eq:semidiscreteDG} results in the semi-discrete form
\begin{align}
  \deriv{\vect{U}_{k}}{t}
  =-\f{1}{w_{k}\sqrt{\gamma}_{k}\dx}
  \Big\{
  \Big[\,
    \sqrt{\gamma}\widehat{\vect{F}}\ell_{k}\big|_{\xH}
    -\sqrt{\gamma}\widehat{\vect{F}}\ell_{k}\big|_{\xL}
  \,\Big]
  -\sum_{q=1}^{N}w_{q}\,\sqrt{\gamma}_{q}\,\vect{F}(\vect{U}_{q})\,\pderiv{\ell_{k}}{\eta}(\eta_{q})
  \Big\} + \vect{S}(\vect{U}_{k}).
  \label{eq:semidiscreteDiscretized}
\end{align}
Eq.~\eqref{eq:semidiscreteDiscretized} comprises a system of ordinary differential equations (ODEs), which are integrated in time with an ODE solver.  
In Section~\ref{sec:numerical} we use the optimal third-order strong stability-preserving Runge-Kutta (SSP-RK3) method from \cite{shuOsher_1988}.

@@

We run this test to $t=1.8$ using $200$ elements.  
In the two upper panels of Figure~\ref{fig:ShuOsher} we plot the density for various values of the troubled-cell indicator threshold $C_{\TCI}$; $0.0$ (full limiting, green), $0.03$ (magenta), $0.3$ (red), and $3.0$ (blue).  
Larger $C_{\TCI}$ implies less limiting.  
The results are compared with a reference solution computed with $2048$ elements.  
With full limiting, the scheme is unable to capture the density variations behind the shock (cf. upper right panel in Figure~\ref{fig:ShuOsher}).  
When $C_{\TCI}=0.03$, the variations are better resolved, but the amplitudes are much reduced relative to the reference solution.  
With $C_{\TCI}=3.0$, the variations are well resolved and the amplitudes are comparable to the reference.  
\begin{figure}[h]
  \centering
  \begin{minipage}{18pc}
    \includegraphics[width=18pc]{./Figures/ShuOsher_Astronum_2018}
  \end{minipage}\hspace{0.5pc}%
  \begin{minipage}{18pc}
    \includegraphics[width=18pc]{./Figures/ShuOsher_Inset_Astronum_2018}
  \end{minipage} \\
  \begin{minipage}{12pc}
    \includegraphics[width=12pc]{./Figures/ShuOsher_TCI_0015_Astronum_2018}
  \end{minipage}\hspace{0.5pc}%
  \begin{minipage}{12pc}
    \includegraphics[width=12pc]{./Figures/ShuOsher_TCI_0150_Astronum_2018}
  \end{minipage}\hspace{0.5pc}
  \begin{minipage}{12pc}
    \includegraphics[width=12pc]{./Figures/ShuOsher_TCI_1500_Astronum_2018}
  \end{minipage}
  \caption{\label{fig:ShuOsher}Results for the Shu-Osher shock tube problem computed with $200$ elements.  The upper panels show the mass density at $t=1.8$ for various values of the troubled-cell indicator threshold $C_{\TCI}$, compared with a high-resolution (2048 elements) reference solution.  In the lower panels we show elements flagged for limiting in the $xt$-plane for various values of $C_{\TCI}$.}
\end{figure}
In the lower panels of Figure~\ref{fig:ShuOsher} we plot locations in the $xt$-plane of elements flagged for limiting by the troubled-cell indicator: $C_{\TCI}=0.03$ (left), $C_{\TCI}=0.3$ (middle), and $C_{\TCI}=3.0$ (right).  
With $C_{\TCI}=0.03$, limiting is triggered in a wide region around the shock and around the peaks of the three leftmost density variations in the upper left panel in Figure~\ref{fig:ShuOsher}.  
With $C_{\TCI}=3.0$, limiting is only triggered in the shock.

\subsubsection{Kelvin-Helmholtz (KH) Instability}
This hydrodynamic instability is of significant astrophysical relevance --- including in the explosion phase of CCSNe --- and may occur in the interface separating fluids in relative motion.  
We compute this problem on the unit square $D=[0,1]\times[0,1]$, imposing periodic boundary conditions, and with initial conditions and single-mode perturbation taken from \cite{mcnally_etal_2012}.  
We use $256^{2}$ elements and evolve until $t=3.0$, well into the nonlinear regime.  
Results are displayed in Figure~\ref{fig:KelvinHelmholtz}.  
We have computed two models, one with $C_{\TCI}=0.06$ (left panels) and one without limiting ($C_{\TCI}\to\infty$; right panels).

@@

This test is of immediate relevance to simulation of CCSNe.  
The initial conditions, described in detail in \cite{blondin_etal_2003}, have been used by many to study the standing accretion shock instability (SASI) and CCSN explosion dynamics.  
We follow closely the description in \cite{blondin_etal_2003} to initialize this test.  
We use spherical polar coordinates, let $D=[0.2,2.0]$, and $\Gamma=4/3$.  
The gravitational potential is given by the point-mass formula with $GM=0.5$.  
A stationary shock is placed at a radius $R_{\shock}=1$.  
Ahead of the shock, the flow is essentially in free-fall towards the shock with a constant Mach number of $100$.  
The mass accretion rate is held fixed at the outer boundary to $\dot{M}=4\pi$.  
Below the shock, $r\in[0.2,1]$, the settling solution is obtained by solving Bernoulli's equation.  
Matter flows through the inner boundary, and the density and pressure in boundary elements are set by extrapolating from $D$ assuming $\rho\propto r^{-3}$ and $E\propto r^{-4}$, while the momentum components are held fixed to their initial values.

\begin{figure}[h]
  \centering
  \begin{minipage}{18pc}
    \includegraphics[width=18pc]{./Figures/SAS_Astronum_2018}
  \end{minipage}\hspace{0.5pc}%
  \begin{minipage}{18pc}
    \includegraphics[width=18pc]{./Figures/SAS_TCI_Astronum_2018}
  \end{minipage} \\
  \begin{minipage}{36pc}
    \includegraphics[width=36pc]{./Figures/SAS_Perturbed_Astronum_2018}
  \end{minipage}
  \caption{\label{fig:SAS}Results from the SAS test using $256$ elements.  In the upper left panel we plot the mass density (blue) and velocity (red) versus radius for the initial condition (dashed lines) and for $t=100$ (solid lines) from an unperturbed run.  In the upper right panel we show elements flagged for limiting in the $rt$-plane.  In the bottom panel we show results from a perturbed model, where we plot the relative deviation of the pressure below the shock; cf. Eq.~\eqref{eq:pressureDeviation}.}
\end{figure}
